Joint training of classification model and similarity model for low-resource text classification in chatbot

### 2020-09-06 chatbot_classification/aaai_2021/paper.tex
Version 2.

  \section{Methodology}
  In  this  section,  we  will  describe our proposed system of similarity model
  fused  with  classification  model,  named  as  SFC, in detail. We propose two
  approaches  to  implement  the  SFC idea, two-stageg SFC, and its more compact
  version called joint-SFC.

@@

  The network   structure   of   two-stage   SFC  is  composed  with  2  separate  stages  in  charge of
  different  jobs. In stage 1, we can use any auxiliary text classification model or tool to provide
  top-K  most  related class labels. These selected class labels will later be used to sample sentence pairs of both positive
  samples  and negative samples for stage 2. Then, in stage 2, we will use sentence pair similarity model to identify the class label with strongest semantic similarity to the user input sentence. Suppose we have a training dataset
  $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$  of  $N$ data points, in which $x_{i}$
  is the user input query and $y_{i}$ is a single class label from a class label
  set  of  $C$  classes  in total. If we want to generate sentence pair dataset
  $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$,   where   $(x,   x')_{j}$  is
  sentence  pair  grouped  by  2  user  input  queries and $l_{j}\in\{0, 1\}$ is
  similarity class label, see Equ. \ref{eq:similarity label},
  \begin{align}
    l_{j}=
    \begin{cases} \label{eq:similarity label}
      0 & y_{x}\neq y_{x'}\\
      1 & y_{x}=y_{x'}
    \end{cases}
  \end{align}
  directly  from  original dataset $\mathcal{D}$ without any sampling strategy,
  then  $\mathcal{D'}$  will  have  $M=N^2$ data points of sentence pairs, which
  might be too time-consuming for the training process of stage 2 since we often
  have  hundreds  of class labels under task-specific chatbot scenario. Although
  we  may  only  have  a  few data points for a single class label, $M$ is still
  quite a big number.

  Therefore,  we adopt the idea of adversarial negative sampling strategy raised
  by   \cite{bamler2020extreme}.  The  key  idea  of this sampling strategy is to
  train  with  negative  samples that are hard to be distinguished from positive
  samples. Now suppose we have a user input query $x_{i}$ and top-K most related
  class   labels   set  $\mathcal{C'}=\{c'_{1},  \dots,  c'_{K}\}$  provided  by
  auxiliary  model  or  tool in stage 1, where $K$ is a hyperparameter set by us
  that  controls  candidate  class  number,  we  apply  the adversarial negative
  sampling  strategy  into our sentence pair sampling process by the following 2
  steps:

  The  first step is \emph{positive sentence pair sampling}. During the training
  process,  we have to first make sure that the ground truth class label $y_{i}$
  for  $x_{i}$  is  within  the  candidate  class  label  set $\mathcal{C'}$. If
  $y_{i}\notin  \mathcal{C'}$,  we will manually add $y_{i}$ into $\mathcal{C'}$
  by  replacing  $c'_{K}$  with  $y_{i}$,  since $c'_{K}$ is the least promising
  candidate  label  according  to the auxiliary model in stage 1. Afterwards, we
  will randomly sample $P$ sentences $\{x'_{0},x'_{1}, \dots, x'_{P}\}$ with the
  same  class  label  as  $y_{i}$  from dataset $\mathcal{D}$ to form the set of
  sentence  pairs  with positive label $\mathcal{P'}_{i}=\{[(x_{i}, x'_{0}), 1],
  [(x_{i},  x'_{1}),  1],  \dots,  [(x_{i},  x'_{P}), 1]\}$, where $P$ is also a
  hyperparameter  set  by  us  that  controls  the number of sentences we should
  sample from each class.

  In  this  way,  we  can generate the sentence pair dataset $\mathcal{D'}$ for
  stage  2 based on the top-K class label set $\mathcal{C'}$ provided by stage 1
  as   $\mathcal{D'}=\{\mathcal{P'}_{i}\cup   \mathcal{N'}_{i}\}_{i=1}^{N}$,  in
  which we will have $M=N\cdot K\cdot P$ data points of sentence pairs in total.

  After  explaining the data sampling principle for sentence pairs and how the 2
  stages  in  SFC  system  connect  with  each other, let's go deeper into model
  structure within these two stages respectively to see how it works.

@@

  As for  stage  1,  our  two-stage SFC  system  can adapt to any auxiliary model or even
  searching  tool  based  on  Term Frequency-Inverse Document Frequency (TF-IDF)
  retrieval.  In  this  paper,  however,  we  choose to fine-tune pretrained model such as RoBERTa on our
  task-specific  dataset $\mathcal{D}$ by adding a softmax classifier on top of
  it for better
  overall  performance.  According  to  our  experimental  result,  RoBERTa  can
  guarantee  that  approximately  95\%  ground  truth labels be included in top-5
  candidate  class  labels  with  highest  probability if it has been fine-tuned on
  task-specific  dataset and can obtain a well performance measured by F-score over
  0.75 on this dataset.  So  we use RoBERTa as the encoder of classification model in our setting at
  stage 1 to ensure low limitation for the performance of stage 2.

  For single-sentence text classification tasks, given a data point $x_{i}$ with
  class  label  $y_{i}$  from  dataset  $\mathcal{D}$, RoBERTa takes the final
  hidden  state  $h_{i}$  of the first token [CLS] as the representation for the
  whole   sequence  of  $x_{i}$.  Let's  suppose  we  have  a  linear  layer  as
  ${\Phi}^C_{i}=W^Ch_{i}+b^C$,  where  $W^C$ and $b^C$ are trainable parameters,
  the    probability    score    for    $x_{i}$    can    be    calculated    as
  $p^C_{i}=softmax(W^Ch_{i}+b^C)=softmax({\Phi}^C_{i})$.   Then,  the  loss  for
  stage 1 classification model is shown in Equ. \ref{eq:classification loss}, in
  which  the  first  term  $\varPhi_{i,y_{i}}^C$ encourages high probability for
  prediction of the correct label $y_{i}$.

@@

  As for  stage 2, we also choose RoBERTa as the encoder for the sentence pair similarity model. Before sending sentence pairs to Roberta, we
  pack  sentence  pairs  into a single sequence and separate them with a special
  token  [SEP].  Roberta  will  also  add  a  learned  embedding  to every token
  indicating  whether  it belongs to the first sentence or the second one. Then,
  we  utilize  the  cross-attention  mechanism  inside RoBERTa, and then extract
  semantic relationship between 2 sentences from the output softmax layer, which
  is placed on top of the last hidden representation of the [CLS] token.

  Before we start to fine-tune our sentence-pair model on task-specific dataset,
  we   first  fine-tune  RoBERTa  on  Quora  dataset \cite{iyer2017first},  which
  contains  404,290  potential  duplicate question pairs, for transfer learning.
  Then   we   continue  to  do  multi-task  training  on  sentence-pair  dataset
  $\mathcal{D'}$ sampled from the top-K candidate class labels provided by stage
  1.

  \subsection{Multitask based training}
  The 2 tasks for stage 2 training process are listed below.

  \subsubsection*{Task 1: Regular sentence-pair similarity task}

  For  sentence  pair semantic similarity task, given a data point $(x, x')_{i}$
  with       similarity       label       $l_{i}$       from       data      set
  $\mathcal{D'}=\{[(x,x')_{i},l_{i}]\}_{i=1}^{M}$,   Roberta   takes  the  final
  hidden  state  $h_{i}$  of the first token [CLS] as the representation for the
  sequence of packed sentence pair $(x, x')_{i}$. Let's suppose we have a linear
  layer  as  ${\Phi}^S_{i}=W^Sh_{i}+b^S$,  where  $W^S$  and $b^S$ are trainable
  parameters,  and  the probability score for $(x, x')_{i}$ can be calculated as
  $p^S_{i}=softmax({\Phi}^S_{i})$.

@@

  \subsubsection*{Task   2:  Classification  task  on  top-K  classes  based  on
  sentence-pair model}

@@

  \subsection{Joint SFC with joint model structure}
  In  two-stage  SFC,  Stage 1 and stage 2 are separate, so they do not interact
  with  each other during training. In this case, the performance of stage 1 can
  limit  the  potential of stage 2, meanwhile, stage 2 also cannot give training
  feedback back to stage 1 for fine-tuning.

@@

  There  are 3 tasks in total during the training process of joint SFC, shown in
  Fig. \ref{fig:framework}, and the overall loss is also the weighted sum of all
  the   task-specific  losses,  shown  in  Equ.  \ref{eq:joint_SFC_loss}.  Here,
  $\alpha_S$,   $\alpha_K$   and   $\alpha_C$   are  the  weights  for  task  1:
  sentence-pair  similarity,  task  2:  top-K classification and task 3: general
  single sentence classification respectively.
  \begin{align}
    \mathcal{L} = \alpha_S \mathcal{L}^S + \alpha_K \mathcal{L}^K + \alpha_C \mathcal{L}^C
    \label{eq:joint_SFC_loss}
  \end{align}

  Finally,  when  doing inference, for each user input query $x$, we will choose
  the  class label that contains the sentence $x'$ with highest similarity score
  $\varPhi$,  after  being packed into a sentence pair $(x, x')$ with query $x$,
  as our final result for prediction.

>>>>

  \section{Methodology}
  In  this section, we describe our proposed system, similarity model fused with
  classification  model (SFC). We propose two approaches to implement this idea,
  two-stageg SFC, and its more unified version joint-SFC.

@@

  Two-stage   SFC  is  composed  with  two  independent stages  in  charge of
  different  jobs.

@@

  In  stage  1,  we  can  use any auxiliary text classification model or tool,
  such as Term Frequency-Inverse Document Frequency (TF-IDF)retrieval, to
  provide  top-K  most  related  class  labels. These selected class labels will
  later  be  used to sample sentence pairs of both positive samples and negative
  samples  for  stage  2.

  In  this  work,  we  choose pretrained models to fine tune on our data $\mathcal{D}$.
  Due to the excellent performance of RoBERTa in this task, compared to other pretrained
  models, we use RoBERTa as the encoder of classification model in our setting.

  In this stage, given a data point $x_{i}$ with
  class  label  $y_{i}$  from  dataset  $\mathcal{D}$, RoBERTa takes the final
  hidden  state  $h_{i}$  of the first token [CLS] as the representation for the
  whole   sequence  of  $x_{i}$.  Let's  suppose  we  have  a  linear  layer  as
  ${\Phi}^C_{i}=W^Ch_{i}+b^C$,  where  $W^C$ and $b^C$ are trainable parameters,
  the    probability    score    for    $x_{i}$    can    be    calculated    as
  $p^C_{i}=softmax(W^Ch_{i}+b^C)=softmax({\Phi}^C_{i})$.   Then,  the  loss  for
  stage 1 classification model is shown in Equ. \ref{eq:classification loss}, in
  which  the  first  term  $\varPhi_{i,y_{i}}^C$ encourages high probability for
  prediction of the correct label $y_{i}$.

@@

  We continue choosing RoBERTa as the main module in the sentence pair
  similarity  model.

  Then, in stage 2, we will use sentence pair similarity
  model  to  identify  the class label with strongest semantic similarity to the
  user    input    sentence.    Suppose    we    have    a    training   dataset
  $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$  of  $N$ data points, in which $x_{i}$
  is the user input query and $y_{i}$ is a single class label from a class label
  set  of  $C$  classes  in  total. If we want to generate sentence pair dataset
  $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$,   where   $(x,   x')_{j}$  is
  sentence  pair  grouped  by  2  user  input  queries and $l_{j}\in\{0, 1\}$
  denotes the similarity label, see Equ. \ref{eq:similarity label},
  \begin{align}
    l_{j}=
    \begin{cases} \label{eq:similarity label}
      0 & y_{x}\neq y_{x'}\\
      1 & y_{x}=y_{x'}
    \end{cases}
  \end{align}
  directly  from  original dataset $\mathcal{D}$ without any sampling strategy,
  then  $\mathcal{D'}$  will  have  $M=N^2$ data points of sentence pairs, which
  might be too time-consuming for the training process of stage 2 since we often
  have  hundreds  of class labels under task-specific chatbot scenario. Although
  we  may  only  have  a  few data points for a single class label, $M$ is still
  prohibitively large.

  Therefore,  we adopt the idea of adversarial negative sampling strategy from
  the work \cite{bamler2020extreme}.  The  key  idea  is to
  train  with  negative  samples that are hard to be distinguished from positive
  samples. Now suppose we have a user input query $x_{i}$ and top-K most related
  class   labels   set  $\mathcal{C'}=\{c'_{1},  \dots,  c'_{K}\}$  provided  by
  auxiliary  model  or  tool in stage 1, where $K$ is a hyperparameter 
  to control  candidate  class  number,  we  apply  the adversarial negative
  sampling  strategy  into our sentence pair sampling process by the following 2
  steps:

  The  first step is \emph{positive sentence pair sampling}. During the training
  process,  we first make sure the ground truth class label $y_{i}$
  for  $x_{i}$  is  within  the  candidate  class  label  set $\mathcal{C'}$. If
  $y_{i}\notin  \mathcal{C'}$,  we manually add $y_{i}$ into $\mathcal{C'}$
  by  replacing  $c'_{K}$  with  $y_{i}$,  since $c'_{K}$ is the least promising
  candidate  label  according  to the auxiliary model in stage 1. Afterwards, we
  will randomly sample $P$ sentences $\{x'_{0},x'_{1}, \dots, x'_{P}\}$ with the
  same  class  label  as  $y_{i}$  from dataset $\mathcal{D}$ to form the set of
  sentence  pairs  with positive label $\mathcal{P'}_{i}=\{[(x_{i}, x'_{0}), 1],
  [(x_{i},  x'_{1}),  1],  \dots,  [(x_{i},  x'_{P}), 1]\}$, where $P$ is also a
  hyperparameter  set  to control  the number of sentences we should
  sample from each class.

  In  this  way,  we  generate the sentence pair dataset $\mathcal{D'}$ for
  stage  2 based on the top-K class label set $\mathcal{C'}$ provided by stage 1
  as   $\mathcal{D'}=\{\mathcal{P'}_{i}\cup   \mathcal{N'}_{i}\}_{i=1}^{N}$,  in
  which we have $M=N\cdot K\cdot P$ data points of sentence pairs in total.

  Before starting to fine-tune our sentence-pair model on task-specific dataset,
  we   first  fine-tune  RoBERTa  on  Quora  dataset \cite{iyer2017first},  which
  contains  404,290  potential  duplicate question pairs, for transfer learning.
  Then   we   continue  to  do  multi-task  training  on  sentence-pair  dataset
  $\mathcal{D'}$ sampled from the top-K candidate class labels provided by stage
  1.

  \subsubsection*{Multi-task based training}
  The 2 tasks for stage 2 training process are listed below.

  The first task is \emph{regular sentence-pair similarity task}.
  For  sentence  pair semantic similarity task, given a data point $(x, x')_{i}$
  with       similarity       label       $l_{i}$       from       data      set
  $\mathcal{D'}=\{[(x,x')_{i},l_{i}]\}_{i=1}^{M}$,   Roberta   takes  the  final
  hidden  state  $h_{i}$  of the first token [CLS] as the representation for the
  sequence of packed sentence pair $(x, x')_{i}$. Let's suppose we have a linear
  layer  as  ${\Phi}^S_{i}=W^Sh_{i}+b^S$,  where  $W^S$  and $b^S$ are trainable
  parameters,  and  the probability score for $(x, x')_{i}$ can be calculated as
  $p^S_{i}=softmax({\Phi}^S_{i})$.

@@

  The second task is \emph{classification  task  on  top-K  classes  based  on
  sentence-pair model}.

@@

  \subsection{Joint SFC}
  In  two-stage  SFC,  Stage 1 and stage 2 are separate that there is no deep
  interaction
  with  each other during training. In this case, the performance of stage 1 can
  limit  the  potential of stage 2, meanwhile, stage 2 also cannot give training
  feedback back to stage 1 for fine-tuning.

@@

  There  are 3 tasks in total during the training process of joint SFC, shown in
  Fig. \ref{fig:framework}, and the overall loss is also the weighted sum of all
  the   task-specific  losses,  shown  in  Equ.  \ref{eq:joint_SFC_loss}.  Here,
  $\alpha_S$,   $\alpha_K$   and   $\alpha_C$   represent the  weights  for  task  1,
  sentence-pair  similarity,  task  2,  top-K classification and task 3, general
  single sentence classification respectively.
  \begin{align}
    \mathcal{L} = \alpha_S \mathcal{L}^S + \alpha_K \mathcal{L}^K + \alpha_C \mathcal{L}^C
    \label{eq:joint_SFC_loss}
  \end{align}
